Approximation as Differentiation

### 2025-05-24 main/category-of-families.tex
Some simplification of auto-diff.

For any map $f: \RR^m \to \RR^n \times (\RR^m \linearto \RR^n)$ it is useful to define $f_1 = \pi_1 \comp f:
\RR^m \to \RR^n$ and $f_2 = \pi_2 \comp f: \RR^m \to (\RR^m \linearto \RR^n)$.

\begin{definition}[$\Diff$]
\label{def:auto-diff:Diff}
\todo{Overkill -- try and give a construction that lives in $\Fam(\FinVect)$ directly} Define $\Diff$
to be the category where the objects are all finite vector spaces $\RR^n$ and the morphisms are all maps $f:
\RR^m \to \RR^n \times (\RR^m \linearto \RR^n)$ with $f_1$ differentiable. Composition in $\Diff$ is given by:
\begin{align*}
(g \comp f)_1(x) &= f_1(g_1(x)) \\
(g \comp f)_2(x) &= g_2(f_1(x)) \comp f_2(x)
\end{align*}
\end{definition}

We can verify that $\tangents$ is indeed a functor, regardless of whether the linear maps $\RR^m \linearto
\RR^n$ happen to be derivatives. But if $f_2(x)$ is the derivative of $f_1$ at $x$ for any $x \in \RR^m$ and
similarly for $g$ then the derivatives compose according to the forward chain rule and $(g \comp f)_2(x)$ is
the derivative of $(g \comp f)_1$ at $x$.

\subsubsection{Embedding into $\Fam(\FinVect)$}
\label{sec:galois-slicing-auto-diff-via-fam:auto-diff}

Consider the functor $\Diff \hookrightarrow \Fam(\FinVect)$ which send objects $\RR^n$ to constant families of
tangent spaces $(\RR^n, \const(\RR^n))$ and morphisms $f: \RR^m \to \RR^n \times (\RR^m \linearto \RR^n)$ to
pairs $(f_1, f_2)$ of differentiable functions $f_1: \RR^m \to \RR^n$ and families of linear maps $f_2:
\{{f_2}_x: \RR^m \linearto \RR^n\}_{x \in \RR^m}$. This exhibits $\Diff$ as equivalent to a full subcategory
of $\Fam(\FinVect)$.

In that subcategory, for any $(f, \partial f): (\RR^m, \const(\RR^m)) \to (\RR^n, \const(\RR^n))$ and $(g,
\partial g): (\RR^n, \const(\RR^n)) \to (\RR^k, \const(\RR^k))$, we have that for any $x \in \RR^m$:
\begin{itemize}
\item $\partial f_x: \RR^m \to \RR^n$;
\item $\reindex{\partial g}{f}_x = \partial g_{f(x)}: \RR^n \to \RR^k$.
\end{itemize}

>>>>

For any map $f: \RR^m \to \RR^n \times (\RR^m \linearto \RR^n)$ it is useful to define $f_1 = \pi_1 \comp f:
\RR^m \to \RR^n$ and $f_2 = \pi_2 \comp f: \RR^m \to (\RR^m \linearto \RR^n)$. Then composition of such maps can be expressed as:
\begin{align*}
(g \comp f)_1(x) &= f_1(g_1(x)) \\
(g \comp f)_2(x) &= g_2(f_1(x)) \comp f_2(x)
\end{align*}

We can verify that $\tangents$ is indeed a functor, regardless of whether the linear maps $\RR^m \linearto
\RR^n$ happen to be derivatives. But if $f_2(x)$ is the derivative of $f_1$ at $x$ for any $x \in \RR^m$ and
similarly for $g$ then the derivatives compose according to the forward chain rule and $(g \comp f)_2(x)$ is
the derivative of $(g \comp f)_1$ at $x$. We can see these maps as living in $\Fam(\FinVect)$ if we regard
every $\RR^n$ as paired with a constant family of tangent spaces also of the form $\RR^n$ and each map $f:
\RR^m \to \RR^n \times (\RR^m \linearto \RR^n)$ to pairs $(f_1, f_2)$ as the pair of a differentiable function
$f_1: \RR^m \to \RR^n$ and a family of linear maps $f_2: \RR^m \to \RR^m \linearto \RR^n$.

Then for any $(f, \partial f): (\RR^m, \const(\RR^m)) \to (\RR^n, \const(\RR^n))$ and $(g,
\partial g): (\RR^n, \const(\RR^n)) \to (\RR^k, \const(\RR^k))$, we have that for any $x \in \RR^m$:
\begin{itemize}
\item $\partial f_x: \RR^m \to \RR^n$;
\item $\reindex{\partial g}{f}_x = \partial g_{f(x)}: \RR^n \to \RR^k$.
\end{itemize}

### 2025-05-24 main/formal-structures.tex
Some simplification of auto-diff.

\noindent The trivial 1-point lattice, which is both terminal and initial and we write as $0$, is the unit for
$\biprod$. Such a structure $(\biprod, 0)$ where $\biprod$ acts as both a product and a coproduct is called a
\emph{biproduct} (\secref{biproducts}). The presence of a zero object in a Cartesian closed category $\cat{C}$
renders $\cat{C}$ trivial; if $0 \times X \iso X$ then $\Hom{\cat{C}}{X}{Y} \iso \Hom{\cat{C}}{0}{X
\Rightarrow Y} \iso 1$, and $\cat{C}$ has only a single object, up to isomorphism. However although the
biproduct structure rules out Cartesian closure, it will prove crucial for establishing a Cartesian closed
setting in \secref{higher-order}; but let us first see how to develop a suitable categorical setting for
first-order programs.

>>>>

\noindent The trivial 1-point lattice, which is both terminal and initial and we write as $0$, is the unit for
$\biprod$. Such a structure $(\biprod, 0)$ where $\biprod$ acts as both a product and a coproduct is called a
\emph{biproduct} (\secref{biproducts}). The presence of a zero object in a Cartesian closed category $\cat{C}$
renders $\cat{C}$ trivial; if $0 \times X \iso X$ then $\Hom{\cat{C}}{X}{Y} \iso \Hom{\cat{C}}{0}{X
\Rightarrow Y} \iso 1$, and $\cat{C}$ has only a single object, up to isomorphism. However although biproducts
rule out Cartesian closure, they will prove crucial for establishing a Cartesian closed setting in
\secref{higher-order}; but let us first see how to develop a suitable categorical setting for first-order
programs.

### 2025-05-24 main/higher-order.tex
Expand why LatGal isn't CC.
Some simplification of auto-diff.

\section{Interpreting Higher-Order Programs}

\todo{
\begin{enumerate}
\item conditions under which $\Fam(\cat{C})$ is Cartesian closed
\item $\PSh(\cat{C})$ has all small products (inherited from $\Set$)
\item why we need $\PSh_{\CMon}(\cat{C})$
\end{enumerate}
}

>>>>

\section{Interpreting Higher-Order Programs}
\label{sec:higher-order}

\todo{
\begin{enumerate}
\item if $\cat{C}$ has biproducts and all small products then $\Fam(\cat{C})$ is Cartesian closed
\item $\PSh(\cat{C})$ has all small products (inherited from $\Set$)
\item $\FinVect$ and $\LatGal$ have biproducts but not all small products
\item but actually we need $\PSh_{\CMon}(\cat{C})$ rather than $\PSh(\cat{C})$ because $\CMon$-enriched Yoneda embedding preserves biproducts
\end{enumerate}
}
